Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Small fixes:
 * flat -> user defined
 * resolved some notes

\subsection*{Flat contracts}
\label{sec:flat-contracts}

\unsure{Should we define contracts prior to this?}
User defined contracts, also called flat contracts, are a particular
kind of contract that behaves similar to an assertion, and whose
main point is to apply some arbitrary predicate (as arbitrary as the
host language allows it) to a value; if the predicate evaluates to
true, the contract passes, otherwise it fails.

For instance, the contract \nickel{Even} shown on Figure
\ref{fig:contract-higher-order} is a flat contract, defined
by taking the value that needs to be checked, and validating
that is both a number and that the remainder of dividing it by 2
is 0.
And then, as seen on the same figure, we can use it to check that a function always
returns an \nickel{Int} when applied to an \nickel{Even}.

@@

Or even more interesting, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
by applying the function and checking its results,
and then use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:checked-not}.

@@

Finally, the Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.

Even if flat contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are accustomed to express complex predicates
over values.

@@

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
unions and intersections can make \ref{eq:cbn-expansion} \resolved{it} unsound
\emph{even in a pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}
\todo{Since nickel uses | for contract application, I'd move unions
to some other symbol.
I think the best option is to move to $\cap / \cup$}

@@

On the other hand, in dynamically typed languages, it is customary to directly
store different kinds of data in one variable without further ceremony. The
runtime system needs to attach type tags to values anyway, which can usually be
observed using functions like JavaScript's \verb+typeof+. Using such bare
unions\resolved{If you can observe with typeof, then it's pretty much a tagged
union, though the tag is implicit} has already been an idiom in dynamically
typed language for a long time. Being able to type and handle these unions is an
important aspect of the migration to a gradual type system. Indeed, a lot of
\unsure{This is the first time discussing gradual type systems, maybe it should
be introduced earlier, in particular how they relate to contracts
(I think is mentioned on the appendix)}
gradually typed languages either existed originally as dynamically typed
language or aim at backward compatibility with a dynamically typed base
language. This is why despite the presence of a static type system, gradually
typed languages tend to favor bare unions types which is the typed counterpart
of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

@@

\subsection*{Flat contracts do not intersect}
\label{sec:flat-and-inter}

@@

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check the contract \nickel{C}, by
checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g | (String -> String) @& C
      = (fun x => x) 
\end{lstlisting}
\caption{Intersection and flat contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between flat contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch, and vice versa.

@@

Williams, Morris, and Wadler, take a different approach on flat contracts,
they are completely forbidden in their system, explicitly to not encounter
the previously mentioned issue.
This restriction goes against what is usually expected from dynamic checks:
validation of data, and not just inclusion on types.

>>>>

\subsection*{User defined contracts}
\label{sec:flat-contracts}

User defined contracts, sometimes called flat contracts, are a particular
kind of contract that behaves similar to an assertion, and whose
main utility is to apply some arbitrary predicate (as arbitrary as the
host language allows it) to a value; if the predicate evaluates to
true, the contract passes, otherwise it fails.

For instance, the contract \nickel{Even} shown on Figure
\ref{fig:contract-higher-order} is a user defined contract, defined
by taking the value that needs to be checked, and validating
that is both a number and that the remainder of dividing it by 2
is 0.
And then, as seen on the same figure, we can use it to check that a function always
returns an \nickel{Int} when applied to an \nickel{Even}.

@@

Even more interesting, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
by applying the function exhaustively and checking its results,
and then use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:checked-not}.

@@

The Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.

Even if user defined contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are accustomed to express complex predicates
over values.

@@

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
unions and intersections can make \ref{eq:cbn-expansion} unsound
\emph{even in a pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}
\resolved{Since nickel uses | for contract application, I'd move unions
to some other symbol.
I think the best option is to move to $\cap / \cup$}

@@

On the other hand, in dynamically typed languages, it is customary to directly
store different kinds of data in one variable without further ceremony. The
runtime system needs to attach type tags to values anyway, which can usually be
observed using functions like JavaScript's \verb+typeof+. Using such bare
unions\resolved{If you can observe with typeof, then it's pretty much a tagged
union, though the tag is implicit} has already been an idiom in dynamically
typed language for a long time. Being able to type and handle these unions is an
important aspect of the migration to a gradual type system. Indeed, a lot of
\resolved{This is the first time discussing gradual type systems, maybe it should
be introduced earlier, in particular how they relate to contracts
(I think is mentioned on the appendix)}
gradually typed languages either existed originally as dynamically typed
language or aim at backward compatibility with a dynamically typed base
language. This is why despite the presence of a static type system, gradually
typed languages tend to favor bare unions types which is the typed counterpart
of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

@@

\subsection*{User defined contracts do not intersect}
\label{sec:flat-and-inter}

@@

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with user defined contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check the contract \nickel{C}, by
checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g | (String -> String) @& C
      = (fun x => x) 
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that user defined contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between user defined contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch, and vice versa.

@@

Williams, Morris, and Wadler, take a different approach on user defined contracts,
they are completely forbidden in their system, explicitly to not encounter
the previously mentioned issue.
This restriction goes against what is usually expected from dynamic checks:
validation of data, and not just inclusion on types.
